Characterising cheque dishonour cases in India: Causes for delays and policy implications

### 2022-04-14 DOC/paper_chequeDishonour.tex
Removed comments. Made some minor corrections

\usepackage{colortbl}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[parfill]{parskip}
\usepackage[title,titletoc]{appendix}
\usepackage{amsmath}
\usepackage{fnpos}
\usepackage{geometry}
\usepackage{csquotes}
\usepackage{tikz}
\usepackage{pdflscape}

\graphicspath{{../DOC/SLIDES/IMAGES/}}

@@

\usepackage[acronym,nonumberlist,nomain]{glossaries}
\newglossaryentry{adr}{name={ADR},description={Alternative Dispute Resolution}}
\newglossaryentry{crpc}{name={CrPC},description={Code of Criminal Procedure}}
\newacronym{coe}{CoE}{Committee of Experts}
\newacronym{lci}{LCI}{Law Commission of India}
\newacronym{ni}{NI Act}{Negotiable Instruments Act, 1881}
\newacronym{njdg}{NJDG}{National Judicial Data Grid}
\newacronym{sci}{SCI}{Supreme Court of India}
\newacronym{ut}{UT}{Union Territory}

@@

\usepackage[hidelinks,breaklinks,colorlinks=true,allcolors=blue,citecolor = blue]{hyperref}
\usepackage{cleveref}

\title{Characterising cheque dishonour cases in India: Causes for delays and policy implications}
\author{Devendra Damle\thanks{Devendra Damle is a researcher at the National Institute of Public Finance and Policy.} \and Jitender Madaan\thanks{Jitender Madaan is an Associate Professor at the Indian Institute of Technology, Delhi.} \and Karan Gulati\thanks{Karan Gulati is a researcher at the Vidhi Centre for Legal Policy.} \and Manish Kumar Singh\thanks{Manish Kumar Singh is an Assistant Professor at the Indian Institute of Technology, Roorkee.} \and Nikhil Borwankar\thanks{Nikhil Borwankar is a practising advocate.}}

@@

\begin{abstract}

This paper analyses the determinants of case disposition time under the Negotiable Instruments Act, 1881 by creating a unique dataset of roughly 50,000 case level records spread across ten major States in India. Unlike previous studies, this analysis focuses mainly on case characteristics directly linked with procedural matters to estimate the case duration. Regression results show that multiplicity of proceedings, non-appearance of accused, trials being run as summons trial, jurisdictional issues, and reference to mediation significantly increase the case duration. These findings are robust even when considering the number of hearings as a proxy for case disposition time. Additionally, results indicate that, on average, contested cases have a lower duration. However, it takes roughly three additional hearings to resolve. These estimates also indicate that the 2015 amendment to the Negotiable Instruments Act, 1881, which focused on clarifying the jurisdiction issues, and mandatory centralisation of cases against the same drawer, has reduced the case duration by six months.

\end{abstract}

\textit{Keywords: } Negotiable Instruments Act, cheque bouncing, disposition time, court delay, procedural reforms

\textit{JEL Classification:} K12, K19, K41

@@

\newpage
\section{Introduction}
\label{sec:introduction}

India has a slow judiciary - courts are clogged with large backlogs \autocite{moog1992delays, debroy2008justice, dutta2019modernise}. As of 8th December 2021, over forty-one million cases were pending across district courts \autocite{njdg2021}. It is estimated that judicial delays cost India around 1.5\% of its GDP annually \autocite{dey2016_cost}. Slow judiciaries also have adverse consequences on the structure and efficiency of markets and the quality of life of citizens \autocite{world2004world, chemin2007impact, rao2020institutional}. Therefore, it is hardly surprising that tackling judicial delay has become a top priority for Indian judges and policymakers.

One reason for the large burden on courts is believed to be the large share of \gls{ni} cases.\footnote{In particular, \S~138 of the Negotiable Instruments Act (Dishonour of a cheque for insufficiency, etc., of funds in the account). Act 26 of 1881} While the NI Act contains several provisions that may lead to a dispute, \S~138, in particular, provides that if a cheque (drawn by a person for paying off a debt or a liability) is returned unpaid due to insufficiency of funds or credit, the payer may be imprisoned for up to two years or with a fine up to twice the amount of the cheque, or both. As per the \gls{lci}, they represent 6.5\% and 7.8\% of all institutions and pendency in Indian courts, respectively \autocite{lci2014_arrears}. As per one order of the Supreme Court of India, they reflect more than 15\% of all criminal cases in the District Courts \autocite{sc2020_makwanavstate}. As per another order, they constitute over 30\% of the total pendency in courts.\footcite[Similarly, a study published by the Department of Justice briefly touches on the burden of such cases on the judiciary and posits that they constitute 34\% of pending criminal cases in Maharashtra.][]{sc2020_138, mahadik2018_maharashtra}

Given this increasing volume and persistent mention of \gls{ni} cases as a major cause of judicial delay, \cite{sridhar2017_cheque} examined the behaviour of cheque dishonour cases in India by looking at the DAKSH database of 67,000 cases spread across Indian courts. In most cases, she found that resolution is delayed well beyond the statutorily prescribed timeline of 180 days. Using 50,000 case level data for eight States and two Union Territories, we also estimate the survival probability (using non-parametric Kaplan-Meier estimation) to understand the proportion of unresolved cases within the stipulated time.\footnote{The dataset used for this analysis is documented in section \ref{sec:data-description}.} Figure \ref{fig:stateSurvival} shows the survival probability (case ongoing) for the \gls{ni} cases for different States. As per our estimate, less than 20\% of the cases get disposed of within the prescribed timeline of 180 days. Also, the probability of case completion within 500 days is less than 50\% for most States.

\begin{figure}[!ht]
\centering
\caption{Survival probability of cases across States}\label{fig:stateSurvival}
\footnotesize
\textit{Note:} The X-axis shows the number of days that a case takes from the date of admission until the outcome. The Y-axis shows the probability of the case continuing up to a given number of days.
\includegraphics[scale=0.35]{survstates.pdf}
\end{figure}

Since cheque dishonour cases are a large burden on the judiciary, various committees, institutions, and government bodies have focused on expediting disposal. For example, the \cite{lci2008_138}, by relying on newspaper reports concerning the proportion of cheque dishonour cases, recommended setting up Fast Track Magisterial Courts (for a detailed discussion, refer to \cite{bhan2015_placing}). The Act was also subject to amendments in 2015 and 2018 to deal with the various procedural matters. Similarly, the Supreme Court has issued directions to lower courts and various financial institutions to take a more pragmatic and realistic approach for the speedy disposal of \gls{ni} cases.

In 2020, the Supreme Court of India took on board a suo-motu case concerning the “expeditious trial of cases under section 138 of the Negotiable Instruments Act 1881” \autocite{sc2020_138}. Among other directions, the court formed a \gls{coe}\footnote{Headed by Hon’ble Mr. Justice RC Chavan, former Judge of the Bombay High Court.} and appointed Amici Curiae\footnote{Mr. Sidharth Luthra (Sr Advocate) and Mr. K Parameshwar (Advocate).} to assist the court, to study processes to expedite the disposal of complaints under \S~138 of the \gls{ni}. In late 2020, the Amici submitted their recommendations. This included (i) increasing the use of pre-and post-summons mediation, (ii) expediting the service of summons to reduce absconsion, (iii) addressing the multiplicity of proceedings, etc. (for additional detail, refer to \cite{amicus2020_submission}).

Given this background, the purpose of this paper is fourfold. First, using case level eCourts data, for the first time in India, we estimate the real judicial caseload due to the \gls{ni}. Second, we estimate the average case disposal time and its variation over time and across States. Third, we build a conceptual framework to establish the link between specific case characteristics and procedural delay. Furthermore, we empirically test our conceptual framework of procedural determinants of case disposal time using regression analysis. In line with \cite{bielen2015}, using case level interim and final orders data, we establish the link between the specific case characteristics and procedural issues while controlling for inherent court and judge characteristics. The empirical literature on the length of court proceedings currently focuses mainly on determinants at the country and court levels, e.g. public budgets for courts, backlogs, number of judges, etc. Some studies also focus on the case level determinants like the number of parties involved, number of pleadings, etc. Given access to the interim and final orders data, we focus on the case characteristics that can be directly targeted using procedural reforms.

Additionally, we examine the role of the Negotiable Instruments (Amendment) Bill, 2015, which was passed by the Parliament in December 2015. The amendment focused on clarifying the jurisdiction-related issues for filing cases for an offence committed under \S~138 of the Negotiable Instruments Act, 1881. It also mandated the centralisation of cases against the same drawer. These reforms were specially drafted to accelerate court proceedings even though no evidence-based insight into the causes of prolonged trials was available. We analyse the effect of the 2015 amendment by comparing the duration of cases handled before and after their implementation using regression analysis. Our research makes it possible to evaluate the effectiveness of these reforms, at least for the States for which we gathered the data.

Our estimates suggest that cheque dishonour cases represent 13.2\% of courts' workload (pending and disposed) in India. This number could be underestimated because, in about 36\% of the cases in the sample, orders were either unavailable or could not be parsed. We also find significant variation in case disposal time across states. Median disposal time varies from as low as 218 days in Karnataka to as high as 547 days in Himachal Pradesh, with a median disposal time of 395 days. We also find that only a fifth of the cases gets disposed of within the prescribed limit of six months.

Controlling for State and year fixed effects, our findings suggest that, all else being equal, cases where the accused fails to appear take an additional 201 days and 7 hearings to dispose of. Similarly, cases run as summons trials take 112 days and 7.4 hearings more than summary trials. Cases with jurisdictional issues take 271 days and 5.5 hearings more to dispose of than cases without these issues. Cases with a multiplicity of proceedings take an additional 168 days and 9.9 hearings to dispose of than those without. Contested cases take fewer days to dispose of than uncontested cases, even though they take more hearings. Our results also indicate that the 2015 amendments have significantly reduced case duration -- reducing the case duration by more than six months.

The rest of this paper is organised as follows: after this introduction, \cref{sec:history} elaborates on the history of cheque dishonour provisions in India and their relation to judicial delays. In \cref{sec:select-case-char}, based on our conversations with practising lawyers, other stakeholders, and the recommendations of the Amici Curiae and the Supreme Court, we figure out the key case characteristics linked to the procedural delay that can be directly targeted by court interventions. \Cref{sec:methodology} describes the methodology, and \cref{sec:results} presents the results thereof. In \cref{sec:2015amend}, we analyse the effectiveness of the 2015 amendments by comparing the duration of cases handled before and after the procedural reforms were implemented. Finally, \cref{sec:conclusion} concludes the paper and presents the way forward.

\section{Cheque dishonour and judicial delays} \label{sec:history}

The \acrlong{ni} was enacted to define the law relating to promissory notes, bills of exchange, and cheques. To increase the culture regarding the use of cheques and enhance the credibility of the instrument, the Act was amended in 1988.\footcite{niAmend1988} A new chapter (\S\S~138 to 142) was incorporated for penalties in the event of dishonour of cheques due to insufficient funds in the account of the drawer of the cheque. It also aimed to include adequate safeguards to prevent harassment of honest drawers. \S~138 provided for the circumstances under which a drawer can be penalised for the dishonour of a cheque (for a detailed discussion of \S~138, refer to Appendix \ref{app:understanding}).\footcite{ind1881_niAct}

However, by 2001, these provisions were thought not to have had the desired effect.\footcite{stdcomm2001_138niAct} Large number of cases were pending across the country, and courts could not dispose of cases in a time-bound manner. The punishment was also thought to be inadequate. A Working Group was constituted the same year to review \S~138 of the \gls{ni} and make recommendations regarding the changes needed to effectively achieve the section's purpose.\footcite{wg2001_138} In light of the recommendations of the Working Group, the government decided to bring further amendments to the Act. Among others, the amendments included:

\begin{enumerate}[label=(\alph*)]
\item increasing the punishment from one year to two years;
\item providing discretion to the court to waive the period of one month for taking cognisance of a case;
\item prescribing the procedure for dispensing with preliminary evidence of the complainant;
\item prescribing the procedure for service of summons via speed post or impanelled private couriers;
\item providing for summary trials; and
\item making the offence compoundable.
\end{enumerate}

The amendments aimed at speedy disposal of cheque dishonour cases through summary trial and making the cases compoundable. The punishment provided under \S~138 also was enhanced from one year to two years. These legislative
reforms aimed to encourage the usage of cheques so that regular business transactions and settlement of liabilities could be ensured. The amendments were considered by the Lok Sabha Standing Committee on Finance, which recommended that given the large burden on courts, the proposed amendments be coupled with the creation of specialised courts for \S~138 cases.\footcite{stdcomm2001_138niAct} However, the subsequent Amending Act did not reference specialised courts.\footcite{niAmend2002}

The \gls{lci} took up the proposal for specialised courts in 2008. According to the Commission, the credibility of the financial sector was facing setbacks due to the large pendency of cheque dishonour cases. Relying on an array of judgments by the Supreme Court of India concerning speedy trials,\footcite{sc1978_khatoon, sc1981_champalal, sc2005_surinder, sc2008_krishna} the Commission recommended the introduction of fast track courts to address cases concerning dishonour of cheques under \S~138 of the Act. However, it did not comment on such courts' number or expected workload.\footcite{lci2008_138} The need for additional courts was re-iterated by the \gls{lci} in 2009.\footcite{lci2009_reforms}

While the \gls{lci} has focused on the need for more courts, the Supreme Court has assisted in implementing the amendments. In 2014, noting that the prime reason for delays in \S~138 case was the absence of the accused, the Court held that the magistrate should adopt a pragmatic and realistic approach such as issuing notices and summons via e-mail or push notification to ensure delivery.\footcite{sc2014_iba} In 2018, the court directed banks to give the details of the e-mail of the accused to the complainant for service through e-mail. The court also directed that: (i) cases must be dealt with summarily, (ii) the evidence of the complainant must be conducted within three months, (iii) an endeavour must be made to conclude the trial within six months, (iv) the trial, as far as practicable, must be held on a day-to-day basis, and (v) High Courts may pass additional localised directions for speedy disposal of cases.\footcite{sc2018_meters}

The Act was also amended in 2015 and 2018 to - clarify the appropriate area of the jurisdiction where cheque dishonour cases can be filed,\footcite{niAmend2015} and grant power to courts to grant interim compensation, respectively.\footcite{niAmend2018} In particular, the \citetitle{niAmend2015} 2015 aimed to address the difficulties the payee or lender faced in filing cases under \S~138 due to a lack of clarity concerning the appropriate forum. It thus inserted specific provisions concerning the jurisdiction for an offence under \S~138, hoping to ensure that a fair trial of cases under \S~138 is conducted keeping in view the parties' interests by clarifying the territorial jurisdiction for trying cases.

On the other hand, the \citetitle{niAmend2018} 2018 aimed to address the delays caused by unscrupulous drawers of dishonoured cheques due to easy filing of appeals and obtaining a stay on proceedings. As a result, the payee would have to spend considerable time and resources in court proceedings to realise the value of the cheque. To discourage frivolous and unnecessary litigation, it inserted a new section to provide that a court trying an offence under \S~138 may order the drawer of the cheque to pay interim compensation to the complainant in a summary trial or a summons case, where he pleads not guilty to the accusation made in the complaint; and in any other case, upon framing of charge.

Lastly, in March 2020, the Supreme Court of India took on board a suo-motu case concerning the “expeditious trial of cases under section 138 of the Negotiable Instruments Act 1881”.\footcite{sc2020_138} As mentioned, the court-appointed Amici Curiae to assist in the matter, which presented its preliminary report to the court in October of the same year. The Amici made several suggestions to expedite the trial, which among others, include:

\begin{enumerate}[label=(\alph*)]
\item Address jurisdictional issues;
\item Address multiplicity of proceedings; and
\item Expedite service of summons to reduce absconsion;
\item Explore setting up specialised courts;
\item Increase the use of pre- and post-summons mediation;
\item Mandate presenting of a plausible defence before conversion from summary to summons trial; and
\item Summon witnesses only when the accused presents a defence.\footcite{amicus2020_submission}
\end{enumerate}

Noticeably, the recommendations of the Amici Curiae were in line with previously accepted reasons for delays in the disposal. While considering the recommendations, the Supreme Court recommended the Government and High Courts take necessary action where possible and held that they should be the subject matter of deliberation by the \gls{coe} appointed in the same case.\footnote{Headed by Hon’ble Mr. Justice RC Chavan, former Judge of the Bombay High Court.}

\section{Conceptual framework and hypotheses}
\label{sec:select-case-char}

The variables considered in the empirical literature on the length of court proceedings can be classified into three broad categories: (1) cross-country studies focused mainly on determinants at the country level (e.g., public budgets for courts, per capita GDP, etc.) (2) within-country variations in court performance focused on court characteristics (e.g. backlogs, number of judges, etc.) and (3) case level studies focused on the case-specific characteristics like the number of parties involved, number of pleadings, etc. However, procedural reforms target features that hinder the operational functioning of the court. These characteristics are difficult to gauge, and empirical literature on these determinants remains scarce, presumably due to a lack of access to case level data. If we look at past recommendations by the \gls{lci}, the Amici Curiae, the Supreme Court, and various other committees and reports, the target of these interventions fall into five broad categories:

\begin{enumerate}
\item how to bring the accused to the court?;
\item how to run the trials (summary or summons trials)?;
\item whether to refer the case to mediation;
\item how to clarify jurisdictional issues?; and
\item how to avoid the multiplicity of proceedings?
\end{enumerate}

They posit that these aspects likely lead to delays. Based on these broad categories, this study examines six unique case characteristics that are the targets of these interventions. The case characteristics whose impact we have chosen to examine were determined based on: (1) the importance of that characteristic and (2) the feasibility of finding reliable information about it in the eCourts data. They are:

\begin{enumerate}
\item the accused fails to appear before the court for at least one hearing;
\item the case is converted to a summons trial;
\item the case is referred to mediation;
\item the case has jurisdictional issues and is transferred to another court, as a result;
\item the case contains a multiplicity of proceedings --- either the dishonoured cheque was issued to satisfy multiple transactions, or multiple cheques were dishonoured; and
\item whether the case was contested.
\end{enumerate}

We discuss the salience of each case characteristic and the accompanying hypotheses below.

\subsection{Non-appearance of the accused}
\label{sec:non-appe-accus}

\textcite{ostrom2000efficiency}, in a study of nine State criminal trial courts in the United States, find that criminal cases where the defendant absconds take 40--90 days longer to dispose of than those where the defendant appears in court.\footnote{The authors attribute this to the fact that in such cases, a separate procedure has to be conducted for the judge to issue a bench warrant to produce the accused before the court.} The defendant failing to appear in court is one of the leading causes of failed hearings \autocite{crownProsecutionService2006_magistrateCourtEfficiency}. Similarly, \textcite{llangasinghe1988_fijiJudicialDelays} found that the accused absconding is the most common reason for delays in case disposal in magistrates' courts in Fiji. The Amici Curiae and the Supreme Court's recommendations are premised on the notion that the non-appearance of the accused leads to delays. We, therefore, seek to examine whether this holds good for cheque dishonour cases in India. We test the following hypotheses:

\begin{description}
\item[H$_{1a}$]: cases, where the accused does not appear for one or more hearings, require more days to dispose
\item[H$_{1b}$]: cases, where the accused does not appear for one or more hearings, require more hearings to dispose
\end{description}

\subsection{Conversion to summons trial}
\label{sec:conv-summ-trial}

We could not find empirical studies on the effectiveness of summary trials in reducing case duration. However, procedural law governing summary trials in the United States and the United Kingdom assumes that summary trials would reduce time delays in case disposal \autocite{miller2003}. At the same time, \textcite{miller2003} argues that regardless of the time savings that may (or may not) result from summary trials, the discretion it gives to courts violates the principle of the right to a trial by jury. Therefore, bearing in mind the Supreme Court and Amici Curiae's recommendation that, to the extent possible, cases should be conducted as summary trials, and they should be converted to summons trials only in specific situations, we test the following hypotheses:

\begin{description}
\item[H$_{2a}$]: cases converted to summons trials require more days to dispose
\item[H$_{2b}$]: cases converted to summons trials require more hearings to dispose
\end{description}

\subsection{Cases referred to mediation} \label{sec:furth-exam-cases}

\textcite{buscaglia1997_latinAmericaCourtDelays} find that, in Latin American courts, when judges mandate mediation, it reduces the overall time required to dispose of a case.\footnote{They conjecture is that the efficiency gains would result in judges having to dedicate less time to adjudicate such cases.} However, a review of empirical research on the time and cost-effectiveness of court-annexed Alternative Dispute Resolution (ADR), including mediation, by \textcite{wissler2004effectiveness} finds no conclusive relationship between the time required to dispose of cases via ADR compared to regular trial. Some studies reviewed by \textcite{wissler2004effectiveness} find that mediation can increase the duration. Some find that it can decrease the duration, while some find no effect. \textcite{heise2010adr} further finds that participation in ADR does not significantly affect case disposal time. However, cases where the parties choose to settle, take less time than ADR. Given the mixed findings on the effect of mediation and the Supreme Court and Amici Curiae's opinion that court-annexed mediation can reduce the time to dispose of cases, we test the following hypotheses:

\begin{description}
\item[H$_{3a}$]: cases referred to mediation require more days to dispose
\item[H$_{3b}$]: cases referred to mediation require more hearings to dispose
\end{description}

\subsection{Jurisdictional issues}

Prima facie, there are four territorial areas involved in a cheque dishonour case - (i) where the issuer ordinarily resides, (ii) where the payee ordinarily resides, (iii) where the cheque was issued, and (iv) where the cheque was presented. Before 2015, the \gls{ni} only specified circumstances under which complaints concerning cheque dishonour could be filed. It did not specify the territorial jurisdiction of the courts where such a complaint had to be filed. This resulted in individuals filing cases in locations not readily accessible to the opposite party. Thus, cases would have to be transferred to suitable courts for hearing. In 2015, the Act was amended to provide that complaints can only be filed in a court in whose jurisdiction the bank branch of the payee lies.\footcite{niAmend2015} While this addressed the lack of clarity in the court where complaints had to be filed; cases commenced in territorial jurisdiction where the accused did not ordinarily reside.\footcite{amicus2020_submission} Such jurisdictional issues, where cases have to be transferred from one court to another, may lead to delays.\footcite{sc2020_138, amicus2020_submission} We thus test the following hypotheses:

\begin{description}
\item[H$_{4a}$]: Cases with jurisdictional issues require more days to dispose
\item[H$_{4b}$]: Cases with jurisdictional issues require more hearings to dispose
\end{description}

\subsection{Multiplicity of proceedings}

\S~219 of the \gls{crpc} provides that when a person is accused of more than one offence of the same kind committed within twelve months, all offences may be tried together, subject to a maximum of three such offences. Similarly, as per \S~220, if a person commits more than one offence in one transaction, she may be charged with all offences and tried at one trial. Experience has shown that a single financial transaction may lead to the dishonour of multiple cheques. However, under the \gls{crpc}, only three offences and, therefore, the dishonour of only three cheques can be tried together. This has resulted in multiple proceedings involving the same issuer (accused) or the same transaction. This increases the case complexity, and reducing such multiplicity may reduce the burden on courts. We thus test the following hypotheses:

\begin{description}
\item[H$_{5a}$]: Cases involving a multiplicity of proceedings require more days to dispose
\item[H$_{5b}$]: Cases involving a multiplicity of proceedings require more hearings to dispose
\end{description}

\subsection{Contested cases}
\label{sec:contested_cases_meth}
When cases are contested, parties lead evidence before a court and rebut arguments. This is likely to lead to longer proceedings since courts have to reach reasoned decisions based on the proceedings. On the other hand, uncontested cases use minimal court resources since parties reach a mutually acceptable decision verified by the court. \textcite{ostrom2000efficiency} find that cases that went to trial took 53--158 days longer to dispose of than cases where the defendant pleaded guilty.\footnote{The variation in the average delay is between the different jurisdictions studied.} \textcite{buscaglia1997_latinAmericaCourtDelays}, in a study of judicial efficiency in Latin American courts, find that willingness to litigate increases the duration of court cases and results in a greater backlog of cases. \textcite{crownProsecutionService2006_magistrateCourtEfficiency} finds that, in magistrate courts in the United Kingdom, contested cases typically take a greater number of hearings to dispose of than uncontested cases. The difference between the time it takes the courts to dispose of uncontested cases and contested cases can yield insights into how efficient the court is at handling cases that go to trial, compared to cases where the defendant is willing to settle or compound. We thus test the following hypotheses:

\begin{description}
\item[H$_{6a}$]: Contested cases require more days to dispose
\item[H$_{6b}$]: Contested cases require more hearings to dispose
\end{description}

\subsection{Summary}

To summarise, these six case characteristics represent different events in a litigation.\footnote{These are not mutually exclusive. One or more of these events can occur in the same case.} We test whether the event's occurrence affects the case duration and the number of hearings required to dispose of it. Table \ref{tab:expected} summarises the expected effects of the identified case characteristics. The potential effectiveness of different stakeholders' recommendations can be indirectly tested through these characteristics.

\begin{table}[!ht]
\caption{Expected effects of case characteristics}
\label{tab:expected}
\centering
\footnotesize
\begin{tabular}{@{}lcc@{}}
\toprule
\multicolumn{1}{c}{\multirow{2}{*}{\textbf{Case characteristic}}} & \multicolumn{2}{c}{\textbf{Expected effect}} \\ \cmidrule(l){2-3}
\multicolumn{1}{c}{} & \textbf{Duration} & \textbf{Hearings} \\ \midrule
Non-appearance of the accused & + & + \\
Conversion to summons trial & + & + \\
Mediation & - & - \\
Jurisdictional issues & + & + \\
Multiplicity of proceedings & + & + \\
Contested cases & + & + \\ \bottomrule
\end{tabular}
\end{table}

\section{Methodology}
\label{sec:methodology}

\subsection{Data description} \label{sec:data-description}

We first drew a random sample of 100,000 cases filed between 1st January 2014 and 31st December 2018 across India from the database published by Development Data Lab.\footcite{devdatalabs2021_eCourtsData} {\color{red} Devendra: Briefly describe DevDataLab data in two lines and why we are using this.} This window was chosen based on consultations with other researchers and practising advocates on the quality and completeness of data. We then attempted to download the case details for these 100,000 cases from the eCourts database.\footcite{ecourts2022} We were able to download case details for approximately 89,000 cases.\footnote{This difference was caused by non-functional links and corrupted or non-machine data.} We used regular expressions to check whether the ``Act Name'' field references the \gls{ni}. This was the first level of case classification. However, from our previous experience with eCourts data, we expected NI Act cases among those where the Act Name field did not match the search parameters.\footcite[The Act Name field sometimes does not contain the name of the substantive Act but rather the name of the procedural Act (e.g., CrPC) or, in rare cases, the subject matter of the case. This is mainly because case classification and nomenclature practices vary from State to State. However, in rare cases, the field is not populated or contains erroneous information. For details, please see,][]{damle2020_ecourtsData} So we used regular expressions to search for references to the \gls{ni}, cheque bounce/dishonour along with \S~138 in the interim and final order texts, where available. We tagged the cases found through this protocol as related to cheque dishonour under \S~138 of the \gls{ni}.

Data availability and quality vary significantly across States. For Arunachal Pradesh, Andaman and Nicobar Islands, and Nagaland, we could not download any information from the eCourts database. For Ladakh, Sikkim, Mizoram, Meghalaya, and Manipur, we do not find a single case referencing the \gls{ni}. For the DNH at Silvasa, we found only one NI Act case. In the case of Puducherry, we did not find any interim or final order for NI Act cases. Given our limitation to processing information recorded in regional language, we dropped States where less than 2\% of the final orders (as a proportion of total NI cases) were in English (Daman and Diu, Uttarakhand, Uttar Pradesh, Rajasthan, Tripura, Chhattisgarh, Jammu and Kashmir, and Bihar). We also dropped eight States where interim or final orders were available in English, but they represented less than 1\% of the cases filed (Telangana, Jharkhand, Kerala, Tamil Nadu, Assam, Orissa, West Bengal, and Madhya Pradesh). Consequently, we could not include all the States of India in our study. Based on the criteria above, of the 29 States (including Jammu and Kashmir) and 7 Union Territories, our final sample covers 8 States and 2 Union Territories, shown in Table \ref{tab:sample_desc}.

\begin{table}[!ht]
\caption{Sample description} \label{tab:sample_desc}
\centering
\footnotesize
\begin{tabular}{lrrr|r}
\toprule
\textbf{State} & \textbf{\gls{ni} cases} & \textbf{Non-\gls{ni} cases} & \textbf{\%\gls{ni} cases} & \textbf{Total cases}\\ \midrule
Andhra Pradesh & 2638 & 18567 & 12.4 & 21205\\
Chandigarh & 731 & 1364 & 34.9 & 2095\\
Delhi & 5202 & 14742 & 26.1 & 19944\\
Goa & 399 & 2713 & 12.8 & 3112\\
Gujarat & 6752 & 49476 & 12.0 & 56228\\
Haryana & 5319 & 33542 & 13.7 & 38861\\
Himachal Pradesh & 1164 & 11948 & 8.9 & 13112\\
Karnataka & 11184 & 75807 & 12.9 & 86991\\
Maharashtra & 8875 & 82673 & 9.7 & 91548\\
Punjab & 5885 & 24739 & 19.2 & 30624\\
\midrule
\textbf{Total} & \textbf{48149} & \textbf{315571} & \textbf{13.2} & \textbf{363720}\\ \bottomrule
\end{tabular}
\end{table}

For these 10 States and Union Territories, we drew another random sample of 500,000 cases filed between 1st January 2014 and 31st December 2018 from the Development Data Labs database.\footcite{devdatalabs2021_eCourtsData} We then attempted to download case details for these 500,000 cases. We were able to successfully download case details for 4,68,855 cases. We removed cases from District and Sessions courts since these courts hear appeals and not the original matters. This gave a sample of 3,63,720 cases. As with the preliminary analysis, we used pattern-matching based on regular expressions to identify cases related to \gls{ni}. In total, our sample has 48,149 \gls{ni} cases. We manually checked the NI Act classification for 1700 cases from the sample of 3,63,720 to check for erroneous classification. We found one false positive and no false negatives.\footnote{The false positive was because, for one case in Haryana, the court had dumped all the judgments issued on that date of disposal in the PDF file containing the order. As a result, the text for an actual NI Act case (along with all other cases disposed of on that day) was erroneously included in the PDF of the judgment of that case. We further checked for similar erroneous data and found one more instance from the same court. However, none of the judgments in the PDF nor the case itself was related to NI Act. So, effectively, it was correctly classified.} However, the data limitation above also limits our ability to reliably check for false negatives. Table \ref{tab:sample_desc} shows a summary of the sample. As shown, check dishonour cases in India represent 13.2\% of courts' workload (pending and disposed). This number could be underestimated because orders cannot be parsed in approximately 36\% of the cases in our sample. Some may be NI Act cases, even though the Act Name field does not say so.

Table \ref{tab:state_disposal} shows the number of disposed and pending cases State-wise and the median time is taken to dispose of cases. According to \S~143 (3) of the \gls{ni}, courts are required to, as far as reasonably possible, complete NI Act cases within six months, i.e., 180 days. However, as the table shows, the median time to dispose of these cases is above the recommended limit in all States. Of the 37,479 disposed of cases in the sample, 10,171 --- i.e., 27\% cases --- were disposed of within the prescribed limit of six months.

\begin{table}[!ht]
\caption{State-wise disposed cases and time to dispose}\label{tab:state_disposal}
\centering
\footnotesize
\begin{tabular}{lrrrr}
\toprule
& & & \textbf{Median days} & \textbf{Median hearings} \\
\textbf{State} & \textbf{Disposed cases} & \textbf{Pending cases} & \textbf{to dispose} & \textbf{to dispose} \\
\midrule
Andhra Pradesh & 2154 & 484 & 451 & 16 \\
Chandigarh & 703 & 28 & 453 & 9 \\
Delhi & 3910 & 1292 & 518 & 8 \\
Goa & 280 & 119 & 352 & 16 \\
Gujarat & 5183 & 1569 & 323 & 13 \\
Haryana & 4241 & 1078 & 488 & 12 \\
Himachal Pradesh & 673 & 491 & 547 & 16 \\
Karnataka & 10221 & 963 & 218 & 6 \\
Maharashtra & 5024 & 3851 & 520 & 19 \\
Punjab & 5080 & 805 & 440 & 13 \\
\midrule
Overall & 37469 & 10680 & 395 & 11 \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\usepackage{longtable}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[parfill]{parskip}
\usepackage[title,titletoc]{appendix}
\usepackage{amsmath}
\usepackage{fnpos}
\usepackage{geometry}
\usepackage{csquotes}
\usepackage{tikz}
\usepackage{pdflscape}

@@

\usepackage[acronym,nonumberlist,nomain]{glossaries}
\makeglossary
\loadglsentries{acronyms.tex}

@@

\usepackage[hidelinks,breaklinks]{hyperref}
\usepackage{cleveref}

\title{Characterising cheque dishonour cases in India: Causes for delays and policy implications}
\author{Devendra Damle\thanks{Devendra Damle is a researcher at the National Institute of Public Finance and Policy.}, Jitender Madaan\thanks{Jitender Madaan is an Associate Professor at the Indian Institute of Technology, Delhi.}, Karan Gulati\thanks{Karan Gulati is a researcher at the Vidhi Centre for Legal Policy.},\\ Manish Kumar Singh\thanks{Manish Kumar Singh is an Assistant Professor at the Indian Institute of Technology, Roorkee.} and Nikhil Borwankar\thanks{Nikhil Borwankar is a practicing advocate.}}

@@

\begin{abstract}
Cheque-dishonour cases constitute a significant portion of litigation in India. These cases are thought to be major contributors to overall pendency and delays in courts. The Supreme Court of India, in 2020, set up a committee and appointed Amici Curiae to formulate recommendations to reduce delays in cheque-dishonour cases. The recommendations target certain aspects of cheque-dishonour cases, however, empirical evidence supporting these recommendations is lacking. In this study we present an empirical analysis of the impact of six major characteristics of cases, which are the targets of the Supreme Court Committee and Amici Curiae's recommendations. We find that many of the recommendations --- more summary trials, clarifying jurisdictions, ensuring presence of the accused, reducing multiplicity of proceedings, and increasing the number of courts --- may indeed reduce the delays in cheque-dishonour cases. However, the recommendation to encourage more mediation is likely to cause more delays instead of reducing them. We thus demonstrate the importance of using empirical analysis to underpin important policy decisions in judicial procedure.
\end{abstract}

\newpage

Copyright Devendra Damle, Jitendar Madaan, Karan Gulati, Manish K Singh \& Nikhil Borwankar.

This working paper is being circulated for discussion and comment purposes. All rights reserved. Short sections of text, not to exceed two paragraphs, may be quoted without explicit permission from the copyright-holders provided that full credit, including copyright notice, is given to the source.

All content reflects the individual views of the authors.

@@

\newpage
\printglossaries

\newpage
\listoftables

\newpage
\section{Introduction}
\label{sec:introduction}
\input{introduction}

\section{Methodology}
\label{sec:methodology}
\input{methodology}

\section{Results}
\label{sec:results}
\input{results}

\section{Conclusion}
\label{sec:conclusion}
\input{conclusion}

\newpage

\section*{Acknowledgements}
The authors would like to thank the DAKSH Centre of Excellence for Law and Technology for their support. The authors are grateful to Sandhya PR and Surya Prakash BS for their extensive feedback on the study. The authors would like to thank Chockalingam Muthian for help with web-scraping. The authors would also like to thank Manaswini Rao and Pramod Rao for reviewing the study and their inputs. We also thank Abhishek Seth for his excellent research assistance. All errors are ours.

\section*{Funding}
This work was supported by the DAKSH Centre of Excellence (CoE) for Law and Technology at IIT Delhi.
